---
title: "Longitudinal project - PBTA Cohort - Copy data file from: /Users/chronia/CHOP/projects/OpenPedCan/OpenPedCan-analysis/data/v12/histologies.csv May 1st, 2023"
author: 'User ID: `r Sys.getenv("USER")`'
date: "`r Sys.Date()`"
output:
  html_document:
    fig_height: 7
    fig_width: 9
    keep_md: yes
    toc: yes
    df_print: paged
params:
  out_path: "/Users/chronia/CHOP/projects/Longitudinal/Analysis/data/plots/"
  data_path: "/Users/chronia/CHOP/projects/Longitudinal/Analysis/data/data_raw/histologies copy.csv" # path to dir with v12
  data_path_cols: "/Users/chronia/CHOP/projects/Longitudinal/Analysis/data/data_raw/PBTA-germline-histology-groups - plot_mapping.csv"
  sample_name: "PBTA"
  subsample: TRUE
---

Longitudinal Project For PBTA Cohort

In this notebook, we will explore how many histologies, how many samples per histology have matched DNA/RNA, how many have 2, 3, 4, samples for the PBTA Cohort, what patients and diagnoses have how many longitudinal samples and are they all paired RNA/DNA or do we have gaps and provide a summary and a table of each set of samples by patient.

Note the colors in the plots do not match the ones of the PBTA cohort yet. ---TO DO---


# Load libraries
```{r setup, include=FALSE}
attach(params) 
library(tidyverse)
library(writexl)
library(ggplot2)
library(magrittr) # needs to be run every time you start R and want to use %>%
library(ggthemes)
library(scales) # For pretty breaks
library(ggpubr)
library(broom)
library(viridis)
library(hrbrthemes)
library(stringi)
library(ggpubr)
library(rstatix)
library(dplyr)
library(RColorBrewer)
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(furrr))

library(reshape2) 
library(reshape) 
library(flextable)

knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.path = file.path(out_path),
                       dev = c("png", "pdf"))

# enable parallelization
#plan("multiprocess", workers = 4) # 10 cpus
plan("multisession", workers = 4) # 10 cpus
options(future.globals.maxSize = 59600 * 1024^3) # for 64 Gb RAM, 119200 * 1024^3 for 128GB

# set out_path and create dir 
out_path <- paste0(out_path, "Plots-", sample_name, "/")
if (!dir.exists(out_path)) {dir.create(out_path)}

# shuffle function:
shuf <- function(df){
  return(df[sample(1:dim(df)[1], dim(df)[1]),])
}

```

## Load data
```{r }
v12 <- read.csv(data_path, stringsAsFactor=FALSE)
# dim(data)
data_path_cols <- read.csv(data_path_cols, stringsAsFactor=FALSE) # Load plotting colors
merge <- merge(v12,data_path_cols,by=c('cancer_group','broad_histology'),all.x=T) #to merge tables

# to discard when add the col palette for PBTA cohort
mypalette <- brewer.pal(12,"Paired") # colorblindFriendly = TRUE
col5 <- colorRampPalette(mypalette)(5)
col8 <- colorRampPalette(mypalette)(8)

```

All tumor samples have matched normal samples. 
We need to filter normal samples out for further analysis. 
We are using "pathology_diagnosis" instead of "sample_type". 
This way we discard normal samples (as shown in "sample_type") in addition to undiagnosed tumor samples.
From "composition" discard Derived Cell Line, PDX

```{r cohort PBTA, echo=TRUE}
data1 <- as.data.frame(merge)
data <- data1  %>%
   filter(cohort == "PBTA",
          !is.na(pathology_diagnosis),
          !composition %in% c("Derived Cell Line", "PDX"),
          !is.na(cancer_group)) # Discard NAs in cancer type  

data %>% count(tumor_descriptor)
#data <- data1 %>% 
#            filter(cohort == "PBTA",
#                   !is.na(pathology_diagnosis),
#                   !composition %in% c("Derived Cell Line", "PDX"),
#                   !is.na(cancer_group)) %>% # Discard NAs in cancer type
#            count(tumor_descriptor)

# Select rows with NA values in cancer_group
# df <- data[is.na(data$cancer_group), ]
# df %>% count(cancer_group)
# df %>% count(tumor_descriptor)

# Save data after filter 
write.csv(data, paste0(out_path, "data_by_cohort_", "PBTA", "_filter.csv"))

# Replicate tumor_descriptor column and rename to tumor_descriptor_sum. This will summarize some longitudinal info.
# data %>% count(tumor_descriptor)

data$tumor_descriptor_sum <- data$tumor_descriptor
data$tumor_descriptor_sum <- ifelse(data$tumor_descriptor_sum=="Progressive Disease Post-Mortem", "Deceased",
                                 ifelse(data$tumor_descriptor_sum=="Initial CNS Tumor", "Diagnosis",
                                 ifelse(data$tumor_descriptor_sum=="Primary Tumor", "Diagnosis", data$tumor_descriptor_sum)))
data <- data %>% filter(tumor_descriptor_sum %in% c("Diagnosis", "Progressive", "Recurrence", "Deceased")) 

data %>% count(tumor_descriptor_sum)

# Save data after filter 
write.csv(data, paste0(out_path, "data_by_cohort_", "PBTA", "_filter_logitudinal.csv"))

data %>% count(cancer_group)
data %>% count(tumor_descriptor)
data %>% count(tumor_descriptor_sum)
data %>% count(composition)
data %>% count(pathology_diagnosis)

count1 <- data %>% 
          group_by(sample_id, cohort_participant_id, Kids_First_Participant_ID,
                 experimental_strategy, sample_type, tumor_descriptor, tumor_descriptor_sum, primary_site,
                 age_at_diagnosis_days, age_last_update_days, age_at_chemo_start, age_at_radiation_start,
                 pathology_diagnosis, cancer_group, broad_histology) %>% 
          dplyr::count(cancer_group) 

write.csv(count1, paste0(out_path, "data_by_cohort_", "PBTA", "_filter_logitudinal_count.csv"))

```


## All assays

```{r plot1}
count1$experimental_strategy <- factor(x = count1$experimental_strategy, levels = c("WGS", "WXS", "RNA-Seq", "Targeted Sequencing", "Methylation"))

p <- ggplot(count1, aes(x=cancer_group, fill=experimental_strategy))+
        geom_bar()+
        scale_fill_manual(values=col5) +
        theme_classic()+
        ylab("Count cancer_group") +
        ggtitle("cancer_group vs experimental_strategy") +
        guides(color=guide_legend(override.aes=list(size=3))) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
        print(p)
ggsave(paste0("Barplot_cancer_group_experimental_strategy_", "PBTA", ".pdf"), plot=p, path=out_path, device="pdf", height=7, width=10)

```

## Paired genomic, transcriptomic assays

```{r plot2}
count <- count1 %>% filter(experimental_strategy %in% c("WGS", "WXS", "RNA-Seq", "Targeted Sequencing")) # Not included:Methylation
count$experimental_strategy <- factor(x = count$experimental_strategy, levels = c("WGS", "WXS", "RNA-Seq", "Targeted Sequencing"))

p <- ggplot(count, aes(x=cancer_group, fill=experimental_strategy))+
        geom_bar()+
        scale_fill_manual(values=col5) + #scale_fill_manual(values=c(data$plot_group_hex)) 
        theme_classic()+
        ylab("Count cancer_group") +
        ggtitle("cancer_group vs experimental_strategy") +
        guides(color=guide_legend(override.aes=list(size=3))) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
        print(p)
ggsave(paste0("Barplot_cancer_group_experimental_strategy_matched_DNA-RNA_", "PBTA", ".pdf"), plot=p, path=out_path, device="pdf", height=7, width=10)
```

## By cancer_group - all disease stages


```{r plot3}
count$tumor_descriptor_sum <- factor(x = count$tumor_descriptor_sum, levels = c("Diagnosis", "Progressive", "Recurrence", "Deceased"))

p <- ggplot(count, aes(x=cancer_group, fill=tumor_descriptor_sum))+
        geom_bar()+
        scale_fill_manual(values=col5) +
        theme_classic()+
        ylab("Count cancer_group") +
        ggtitle("cancer_group vs tumor_descriptor_sum") +
        guides(color=guide_legend(override.aes=list(size=3))) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
        print(p)
ggsave(paste0("Barplot_cancer_group_tumor_descriptor_sum_matched_DNA-RNA_", "PBTA", ".pdf"), plot=p, path=out_path, device="pdf", height=7, width=10)

```

In the following section, we will provide summary table across different disease stages, histologies, and patients samples.

### Number of tumor samples per each assay

First, we’ll examine how many of each type of assay we have (tumors only). This information is stored in the experimental_strategy column.

```{r}
count1 %>%
  group_by(experimental_strategy) %>%
  tally() %>%
  arrange(desc(n)) %>% 
  regulartable() %>%
  fontsize(size = 12, part = "all")

```



### Paired genomic, transcriptomic assays

## Multiple time points: Diagnosis, Progressive, Recurrence, Deceased
We're going to use the `tumor_descriptor_sum` column here with matched genomic and transcriptomic assays from before.

## By cancer_group 
# Disease stage - descriptors pairs

How many tumor samples per cancer_group and disease_stage?

```{r}
disease_stage_cancer_group <- count %>%
  group_by(Kids_First_Participant_ID) %>%
  summarize(disease_stage = paste(sort(unique(tumor_descriptor_sum)),
                                  collapse = ", "),
            cancer_group = paste(sort(unique(cancer_group)),
                                collapse = ", ")) %>%
  group_by(disease_stage, cancer_group) %>%
  tally() %>%
  arrange(desc(n))

disease_stage_cancer_group %>% 
  regulartable() %>%
  fontsize(size = 12, part = "all")

```


What about when Diagnosis is present and paired with another point in time?

```{r}
Diagnosis_paired <- disease_stage_cancer_group  %>%
  filter(disease_stage != "Diagnosis")  %>% 
  filter(grepl("Diagnosis", disease_stage)) %>%
  regulartable() %>%
  fontsize(size = 12, part = "all")
Diagnosis_paired
```



## By Kids_First_Participant_ID
### Disease stage - descriptors pairs

How many tumor samples per disease_stage?

```{r}
disease_stage <- count %>%
  group_by(Kids_First_Participant_ID) %>%
  summarize(disease_stage = paste(sort(unique(tumor_descriptor_sum)),
                                  collapse = ", "),
                                collapse = ", ") %>%
  group_by(disease_stage) %>%
  tally() %>%
  arrange(desc(n))

disease_stage  %>% 
  regulartable() %>%
  fontsize(size = 12, part = "all")

```


What about when Diagnosis is present and paired with another point in time?

```{r}
Diagnosis_paired <- disease_stage  %>%
  filter(disease_stage != "Diagnosis")  %>% 
  filter(grepl("Diagnosis", disease_stage)) %>%
  regulartable() %>%
  fontsize(size = 12, part = "all")
Diagnosis_paired
```


